--- 
title: "MATH0007: Algebra for Joint Honours Students"
author: "Matthew Towers"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: plain
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "A one-term course covering the basics of sets, functions, relations, linear algebra, and group theory."
---

```{r, include=FALSE}
options(tinytex.verbose = TRUE)
```

# Welcome {-}

This is the course text for MATH0007, an introductory algebra module for UCL first year joint honours students.  These notes are split into three parts: Chapter \@ref(sets) on sets, functions, and relations, Chapters \@ref(linalg1) and \@ref(linalg2) on linear algebra, and Chapter \@ref(groups) on group theory.

UCL students can find an exact syllabus, along with problem sheets and suggested reading, on the [module Moodle page](https://moodle.ucl.ac.uk/course/view.php?id=6029).

Take a moment to experiment with the controls in the toolbar at the top of the page.  You can search these notes for a word or phrase (for example, to look up a definition). You can adjust the typeface, the font size, and the background colour to make the text easier to read.  You can also download an offline copy of these notes in various formats.  Please get in touch if you've got any accessibility requirements that aren't met - I'm happy to help.

I hope you enjoy the course.  If you have any questions, comments, or corrections then [email me](mailto:m.towers@ucl.ac.uk).

<!-- This is a _sample_ book written in **Markdown**. You can use anything that Pandoc's Markdown supports, e.g., a math equation $a^2 + b^2 = c^2$. -->

<!-- <!-- $$ these\ don't\ get\ numbered $$ --> 

<!-- <!-- \begin{equation} --> 
<!-- <!--  f(k) = k^2 --> 
<!-- <!--  (\#eq:labml) --> 
<!-- <!-- \end{equation} --> 
<!-- <!-- now use \@ref(eq:label) to refer. --> 

<!-- ```{theorem, name="Fermat's Theorem", label="ferm"} -->
<!-- Here is a theorem. -->
<!-- ``` -->

<!-- Now a ref to Theorem \@ref(thm:ferm). -->

<!-- The **bookdown** package can be installed from CRAN or Github: -->

<!-- ```{r eval=FALSE} -->
<!-- install.packages("bookdown") -->
<!-- # or the development version -->
<!-- # devtools::install_github("rstudio/bookdown") -->
<!-- ``` -->

<!-- Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading `#`. -->

<!-- To compile this example to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): <https://yihui.name/tinytex/>. -->

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
